Litrepl: Literate Paper Processor Promoting Transparency More Than Reproducibility

### 2025-01-02 paper/paper.tex
Update paper

The concept of Literate Programming, initially proposed by Donald Knuth,
revolves around explaining to human beings what we want a computer to do. This
approach is embodied in tools like WEB, CWEB, and NOWEB, which focus on
non-interactive, ahead-of-time compilation. The process includes two primary
modes of operation: weaving, which creates human-readable documentation, and
tangling, which generates machine-executable code. Over time, the concept has
evolved, showing a trend towards simplification \textcite{Knuth1984lp},
\textcite{Ramsey1994lps}.

Concurrently, a concept of human-computer interaction called the
Read-Evaluate-Print Loop or ''REPL'' gained traction, notably within the LISP an
APL communities.\textcite{Spence1975apl, McCarthy1959recfun, Iverson1962apl}

The combination of a command-line interface and a language interpreter enables
incremental and interactive programming, allowing users to directly modify the
interpreter state. By maintaining human involvement in the loop, this approach
is believed to facilitate human thought processes.

A significant milestone in this area became the IPython terminal, which later
evolved into the Jupyter Project. The creators introduced a new document format
called the Notebook, characterized by a series of logical sections of various
types, including text and code, which could directly interact with programming
language interpreters. This interactive communication, akin to REPL style
programming, allows the creation of well-structured documents suitable for
presentations and sharing. The concept underpinning these developments is termed
Literate Computing, which includes goals of spanning a wide audience range,
boosting reproducibility, and fostering collaboration. To achieve these
objectives, several technical decisions were made, notably the introduction of
bidirectional communication—between the computational core, known as the Jupyter
Kernel, and the Notebook acting as a client, along with another layer of
client-server communication between the Notebook web server and the user’s web
browser \textcite{Perez2007IPython, Granger2021litcomp, Kluyver2016jupnb}.

While we recognize the importance of these goals within the Literate Computing
framework, we argue that reproducibility is paramount. Successfully addressing
reproducibility would alone suffice to enhance communication over time among
research project participants and significantly expand the audience.  However,
as it became clear (\textcite{Dolstra2010, Vallet2022}), this challenge extends
far beyond the scope of a human-computer interaction system, and even beyond the
typical boundaries of software distribution management for a particular
programming language. A comprehensive solution to the software deployment
problem should operate at the entire operating system level.

After understanding the scale and complexity of the reproducibility problem, we
suggest to change the focus of human-computer interaction towards simplicity and
transparency. We saw an opportunity to implement a tool named "Litrepl", which
would offer the same familiar REPL-style editing, be compatible with existing
text editors, thus maintaining familiar editing practices, and have a gradually
smaller codebase.

We accomplish this through the following steps. First, we utilize simple
bidirectional text streams for inter-process communication with language
interpreters to evaluate code. Second, we advocate for the reuse of existing
text document formats. In both the Markdown and \Latex evaluators that we have
implemented, simplified parsers are utilized to distinguish code and result
sections from the rest of the document. As of now, we support Python and Shell
interpreter families, as well as a custom AI communication interpreter. Finally,
we strive to leverage POSIX \textcite{POSIX2024} operating system facilities as
much as possible.

>>>>

The concept of \textit{Literate Programming} was formulated by Donald Knuth,
suggesting a shift in focus from writing code to explaining to human beings what
we want a computer to do (\textcite{Knuth1984lp}). This approach is embodied in
the WEB system and its derived family of tools, which translate combined code
and narrative documents into human-readable documentation and machine-executable
code. Over time, the concept has evolved, showing a trend towards simplification
(\textcite{Ramsey1994lps}).

Concurrently, a concept of human-computer interaction often called the
\textit{Read-Evaluate-Print Loop} or ''REPL'' gained traction, notably within
the LISP an APL communities (\textcite{Spence1975apl},
\textcite{McCarthy1959recfun}, \textcite{Iverson1962apl}).

The combination of a command-line interface and a language interpreter enables
incremental and interactive programming, allowing users to directly modify the
interpreter state. By maintaining human involvement in the loop, this approach
is believed to facilitate human thought processes (\textcite{Granger2021litcomp}).

A significant milestone in this area became the IPython interpreter
(\textcite{Perez2007IPython}), which later evolved into the Jupyter Project. The
creators introduced a new document format called the Jupyter Notebook
(\textcite{Kluyver2016jupnb}), characterized by a series of logical sections of
various types, including text and code, which could directly interact with
programming language interpreters. This interactive communication, akin to REPL
style programming, allows the creation of well-structured documents suitable for
presentations and sharing. The concept underpinning these developments is termed
\textit{Literate Computing}\cite{Perez2015blog}, which includes goals of
spanning a wide audience range, boosting reproducibility, and fostering
collaboration. To achieve these objectives, several technical decisions were
made, notably the introduction of bidirectional communication between the
computational core, known as the Jupyter Kernel, and the Notebook web-based
renderer, along with another layer of client-server communication between the
renderer and the user’s web browser.

While we recognize the importance of all goals within the Literate Computing
framework, we argue that the goal of reproducibility is paramount. Successfully
addressing reproducibility would alone suffice to enhance communication over the
time and space separated researchers and significantly expand the audience.
However, as it became clear (\textcite{Dolstra2010}), this challenge extends
beyond the scope of a single human-computer interaction system, and even beyond
the typical boundaries of software distribution management for a particular
programming language. A comprehensive solution to the software deployment
problem should operate at the entire operating system level.

Following \textcite{Vallet2022}, we suggest changing the focus of human-computer
interaction towards simplicity and transparency. We saw an opportunity to
implement a tool that would offer familiar REPL-style editing, be compatible
with existing code editors and text formats, thus maintaining familiar editing
practices, and have a gradually smaller codebase.

We introduce \textit{Litrepl}, a processor designed to achieve these goals
through several means. First, we employ straightforward bidirectional text
streams for inter-process communication with language interpreters to evaluate
code.  Second, we advocate for the reuse of existing text document formats. In
both the Markdown and \Latex evaluators that we have implemented, simplified
parsers are utilized to distinguish code and result sections from the rest of
the document.  As of now, we support Python and Shell interpreter families, as
well as a custom large language model communication interpreter. Finally, we
strive to leverage POSIX\cite{POSIX2024} operating system facilities as much as
possible.
